On mapping errors and seeding heuristics for DNA alignment

### 2019-04-10 MEM.tex
finished correcting typos

The value of $\beta$ has little importance if $p$ is small. Say that
$p=0.01$ and that a read generates a hit in the genome such that 33
nucleotides need to be aligned after seeding. If the hit is random there
is a 99.99\% chance that at least 15 are mismatches. The denominator of
expression (\ref{eq:bayes}) is approximately $1 +
4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
support for the hypothesis that the hit is random is overwhelming.
Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
fewer nucleotides are mismathes. In this case, the denominator is
approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
10^{-9}$, the support for the hypothesis that the hit is not random is
overwhelming. If $p$ is small, we do not need to worry about the value of
$\beta$, one can choose for instance $\beta=1/2$ so that the term
$(1-/\beta)/\beta$ disappears from expression (\ref{eq:bayes}).

@@

The method is general, but it is important to clearly state the
assumptions it depends on. First and most importantly, we have ignored
insertions and deletions. We assume that the sequencing errors are
substitutions only, which makes the method adapted to the Illumina
technology, but not to deletion-prone instruments such as the Oxford
Nanopore technology. We also assume that insertions and deletions never
occur among duplicated sequences. This is obviously incorrect, but our
initial tests with real data suggest that this is a minor impediment (most
likely because this is only one of several simplififying assumptions).
Incorporating insertions and deletions would make the theory intractable,
so it is presently unclear how to deal with this type of error.

>>>>

The value of $\beta$ has little importance if $p$ is small. Say that
$p=0.01$ and that a read generates a hit in the genome such that 33
nucleotides need to be aligned after seeding. If the hit is random there
is a 99.99\% chance that at least 15 are mismatches. The denominator of
expression (\ref{eq:bayes}) is approximately $1 +
4.3\cdot10^{-18}(1-\beta)/\beta$. So unless $\beta < 10^{-17}$, the
support for the hypothesis that the hit is random is overwhelming.
Conversely, if the hit is not random, there is a 99.99\% chance that 4 or
fewer nucleotides are mismathes. In this case, the denominator is
approximately $1+6.8\cdot10^9(1-\beta)/\beta$, so unless $1-\beta <
10^{-9}$, the support for the hypothesis that the hit is not random is
overwhelming. If $p$ is small, we do not need to worry about the value of
$\beta$, one can choose for instance $\beta=1/2$ so that the term
$(1-\beta)/\beta$ disappears from expression (\ref{eq:bayes}).

@@

The method is general, but it is important to clearly state the
assumptions it depends on. First and most importantly, we have ignored
insertions and deletions. We assume that the sequencing errors are
substitutions only, which makes the method adapted to the Illumina
technology, but not to deletion-prone instruments such as the Oxford
Nanopore technology. We also assume that insertions and deletions never
occur among duplicated sequences. This is obviously incorrect, but our
initial tests with real data suggest that this is a minor impediment (most
likely because this is only one of several simplifying assumptions).
Incorporating insertions and deletions would make the theory intractable,
so it is presently unclear how to deal with this type of error.
